---
title: "NLP and NER"
author: "Navdeep Singh"
date: "13/10/2021"
output: html_document
---

# Part 1: Initial setup

## Installing NLP libraries

```{r error=FALSE, eval= FALSE, warning=FALSE, message=FALSE}
# Run only once
install.packages("rJava")
install.packages("openNLP")
install.packages("NLP")
install.packages("tcltk2")
```

## Downloading NLP models

```{r echo=TRUE, eval= FALSE, error=FALSE, warning=FALSE, message=FALSE}
# Run only once
install.packages("openNLPmodels.en", dependencies=TRUE, repos = "http://datacube.wu.ac.at/")
```

## Importing all libraries

```{r echo=TRUE, results='hide', message=FALSE}
#NLP Libraries
library(rJava)
library(openNLP)
library(NLP)

#Tidy data manipulation
library(stringr)
library(dplyr)
library(tidyr)
library(tidytext)
library(readr)
library(stringi)
library(textclean)

#Ingest data
load("saved_data.rda")

#Helper library
library(sqldf)

#Graphics library
library(ggiraphExtra)
library(ggplot2)
library(RColorBrewer)
library(scales)
```

# Part 2: Ingesting corpus

## Importing corpus
```{r import_corpus}
corpus <- bind_rows(hemingway_corpus, fitzgerald_corpus)
```


## Cleaning data

### Stripping extraneous characters

```{r pre_clean_corpus}
corpus_clean <- corpus %>%
  filter(text != "") %>%
  mutate(text = str_replace_all(text, "_", " ")) %>%
  mutate(text = replace_contraction(text)) %>%
  mutate(text = replace_curly_quote(text))
```

## Part 3: Preparing data for NER

### Collapsing content of novels into a string

```{r create_strings}
corpus_text <- corpus_clean %>%
  group_by(title) %>%
  mutate(text = paste(as.character(text), collapse = " ")) %>%
  distinct() %>%
  ungroup()
```

### Converting to nested String object

```{r create_nested_string}
corpus_text_str <- corpus_text %>%
  group_by(title) %>%
  mutate(text = list(as.String(text)))
```

### Setting up NLP pipeline

#### Setting up annotation

```{r initiate_pipeline}
#set pipeline
wordAnnotator <- Maxent_Word_Token_Annotator(language = "en")
sentenceAnnotator <- Maxent_Sent_Token_Annotator(language = "en")
characterAnnotatorEN <- Maxent_Entity_Annotator(language = "en", kind = "person")

pipeline <- list(sentenceAnnotator, wordAnnotator, characterAnnotatorEN)
```

### Initiating annotation

#### Chunking and extracting persons

```{r NER_chunker, message=FALSE}
#create empty df
full_df = as.data.frame(NULL)
chunk_size = 10000

for (j in 1:nrow(corpus_text_str)) {
  #get number of chunks
  chunk <- nchar(corpus_text_str$text[j]) %/% chunk_size
  text <- unlist(corpus_text_str$text[j])
  text <- as.String(text)
  
  #Loop runs through the text section by section and reads each chunk into a df
  
  for (i in 1:chunk) {
    print(paste0(
      "Processing title: ",
      corpus_text_str$title[j],
      " - section ",
      i,
      " of ",
      chunk
    ))
    temp_df = NULL
    
    if (i == 1) {
      m = 1
    }
    
    if (i == chunk) {
      m = n + 1
      n = (nchar(text))
    }
    else{
      n <- m + chunk_size
    }
    
    temp_string = text[m, n]
    
    temp_ann <- NLP::annotate(temp_string, pipeline)
    
    temp_df <-  temp_ann %>%
      as.data.frame %>% 
      filter(type != "word")
    
    temp_df <- temp_df %>%
      mutate(words = str_sub(
        as.character(temp_string),
        start = temp_df$start,
        end = temp_df$end
      )) %>%
      unnest_wider(features)
    
    temp_df <- temp_df %>%
      mutate(author = corpus_text_str$author[j], title = corpus_text_str$title[j]) 
    
    # Stitching together all data
    full_df <- full_df %>%
      bind_rows(temp_df)
    
    m <- m + chunk_size
  }
}

```

#### Creating backup variable

```{r backup_annotated_data}
full_df_backup <- full_df
write_csv(full_df, "full_df_backup.csv")
```


#### Rough cleaning the data

```{r rough_clean}
full_df <-  full_df %>%
  mutate(words = str_remove_all(words, '[:punct:]'))
```

#### Realigning the columns

```{r reshape_df}
full_df <- full_df %>%
  relocate(c("author", "title"), .before = 1) %>%
  select(-id,-constituents) 
```

#### Backing up processed data

```{r backup_annotations, results='hide'}
write_csv(full_df, "annotation_backup.csv") 
```


#### Restoring data from backup (Optional)

```{r restore_annotated_data}
#full_df <- read_csv("annotation_backup.csv")
```

## Part 4: Cleaning annotated text

### Appending locations in each sentence

```{r split_sentence_entity}
df1 <- full_df %>%
  filter(type == "sentence") %>%
  mutate(sentence_num = row_number()) %>%
  select(author, title, words, sentence_num) %>%
  rename(sentence = words)

df2 <-  full_df %>%
  filter(type == "entity") %>%
  mutate(record = row_number()) %>%
  select(title, words, kind)
```

### Cleaning entities thoroughly

```{r computational_cleaning}
df2 <- df2 %>%
  mutate(words = str_replace_all(words, "Dear ", "")) %>%
  mutate(words = str_replace_all(words, "Young ", "")) %>%
  mutate(words = str_replace_all(words, "Ah", "")) %>%
  mutate(words = str_replace_all(words, "Oh", "")) %>%
  mutate(words = str_trim(words, side = "both")) %>%
  mutate(words = str_trim(gsub("[A-Z]{2,}", "", words))) %>%
  mutate(words = str_squish(words)) %>%
  mutate_all( ~ ifelse(. %in% c("N/A", "null", ""), NA, .)) %>%
  drop_na() %>%
  dplyr::filter(nchar(words) > 2) %>%
  distinct()

capital_stopwords <-
  as.data.frame(str_to_title(stop_words$word)) %>%
  rename(words = 1)

df2 <- df2 %>%
  anti_join(capital_stopwords)
```

### Manually cleaning entities

#### Exporting the table of entities

```{r backup_join}
write_csv(df2, "pre_join_entities_new.csv")
```

#### Importing the cleaned table of entities

```{r load_cleaned_data}
#Prep the data
pre_join <- read_csv("pre_join_entities_clean.csv", na = "NA")
```

#### Removing duplicates

```{r limit_join_words}
pre_join <- pre_join %>%
  select(words, kind) %>%
  drop_na()  %>%
  distinct()
```

### Joining entities to sentences

```{r match_dataframes}
#Execute a SQL query
full_join_df <- sqldf("SELECT *
      FROM df1
      LEFT JOIN pre_join ON df1.sentence LIKE '%' || pre_join.words || '%'")

full_join_df <- full_join_df %>%
  distinct()
```

## Part 5: Sentiment analysis

### Cleaning entities + sentences table

```{r exporting_entities_raw}
write_csv(full_join_df, "entities_raw.csv")
```


```{r importing_entities_clean}
clean_entities <- read.csv("entities_clean.csv")
```

### Attaching sentiments to entities

#### Unnesting sentences

```{r unnest}
entities_unnest <- clean_entities %>%
  unnest_tokens(word, sentence)
```

#### Dropping Stopwords

```{r remove_stopwords}
entities_unnest <- entities_unnest %>%
  anti_join(stop_words)
```

#### Attaching NRC sentiments

```{r message=FALSE}
#create sentiment table
entities_sentiment <- entities_unnest %>%
  group_by(author, title) %>%
  inner_join(get_sentiments("nrc")) %>%
  count(sentence_num, sentiment) %>%
  spread(sentiment, n, fill = 0) %>%
  mutate(sentiment = positive - negative)
```

#### Combining sentiments with entitiees

```{r message=FALSE}
entities_matches_sentiment <- entities_unnest %>%
  inner_join(entities_sentiment) %>%
  distinct_at(vars(-word))  
```

#### Creating sentiment table

```{r message=FALSE}
ner_total_sentiment <- entities_matches_sentiment %>% drop_na(words) %>%
  group_by(author, title, words, kind) %>%
  summarise(total = mean(sentiment))  
```

## Part 6: Visualization

### Chart of the top 10 positive entities

```{r message=FALSE}
vis_ner_1 <- ner_total_sentiment %>%
  group_by(title) %>%
  filter(kind == "person") %>% filter(title != "The Old Man and the Sea") %>%
  top_n(5) %>%
  mutate(words = reorder(words, total)) %>%
  ggplot(aes(words, y = total, fill = title, text = paste("Character: ", words, "\n", "% of positive emotion: ", total, sep = ""))) +
  geom_col()  +
  facet_wrap(~ title, scales = "free", ncol = 2) +coord_flip() + labs(x = "Names of characters", 
       y = "% of positive emotion", fill = "Novel", title = "% of Positive Emotions Detected")

vis_ner_1_int <- ggplotly(vis_ner_1, tooltip = "text")
vis_ner_1_int 
```
theme(plot.margin = margin(0, 0, 0, 0, "cm"))
### Chart of the top 10 negative entities

```{r message=FALSE}
vis_ner_2 <- ner_total_sentiment %>%
  group_by(title) %>%
  filter(kind == "person") %>%
  top_n(-5) %>%
  mutate(words = reorder(words, (desc(total)))) %>%
  ggplot(aes(words, x = total, fill = title, text = paste("Character: ", words, "\n", "% of negative emotion: ", 0 - total, sep = ""))) +
  geom_col() +
  facet_wrap(~ title, scales = "free") + labs(x = "Names of characters", 
       y = "% of negative emotion", fill = "Novel", title = "% of Negative Emotions Detected")
vis_ner_2_int <- ggplotly(vis_ner_2, tooltip = "text")
vis_ner_2_int

```


### Visualising entities with NRC emotions

#### Ranking by "total" emotion

```{r radar_plot_total, message=FALSE}
radar_facet <- entities_matches_sentiment %>% drop_na(words) %>%
  select(-positive, -negative, -sentiment) %>% #drop out the unnecessary columns 
  group_by(title, words, kind) %>%
  summarise(across(anger:trust, sum)) %>%
  mutate(total = rowSums(across(where(is.numeric))))  %>%
  arrange(desc(total)) %>%
  head(10)  %>% #Change number to include more or fewer entities
  mutate(across(anger:trust, .fns = ~ round((. / total) * 100))) %>%
  select(-total,-kind)

ggRadar(
  data = radar_facet,
  mapping = aes(color = title, facet = words),
  rescale = FALSE,
  interactive = TRUE,
  use.label = TRUE,
  size = 1,
  legend.position = "right"
)
```


#### Radar plot by the highest positive or negative emotion

```{r message=FALSE}
radar_facet_sentiment <- entities_matches_sentiment %>%

  filter(kind == "person") %>%
  group_by(title, words, kind) %>%
  summarise(across(anger:sentiment, sum)) %>%
  arrange(desc(sentiment))  %>%
  head(10)  %>% #Change number to include more or fewer entities
  select(-positive, -negative, -sentiment, -kind) %>% mutate(total = rowSums(across(where(is.numeric)))) %>% mutate(across(anger:trust, .fns = ~ round((. / total) * 100)))

sentiment_Dick_Diver <- entities_matches_sentiment %>% filter(title == "Tender is the Night") %>%
  filter(words == "Dick") %>%
  group_by(title, words, kind) %>%
  summarise(across(anger:sentiment, sum)) %>% select(-positive, -negative, -sentiment, -kind)

sentiment_Gatsby <- entities_matches_sentiment %>%
  filter(words == "Gatsby") %>%
  group_by(title, words, kind) %>%
  summarise(across(anger:sentiment, sum)) %>% filter(title == "The Great Gatsby") %>% select(-positive, -negative, -sentiment, -kind)

sentiment_Anthony <- entities_matches_sentiment %>% filter(words == "Anthony") %>% filter(title == "The Beautiful and Damned") %>%
  group_by(title, words, kind) %>%
  summarise(across(anger:sentiment, sum)) %>% select(-positive, -negative, -sentiment, -kind)

sentiment_Amory <- entities_matches_sentiment %>% filter(words == "Amory") %>% filter(title == "This Side of Paradise") %>%
  group_by(title, words, kind) %>%
  summarise(across(anger:sentiment, sum)) %>% select(-positive, -negative, -sentiment, -kind)

all_protag_fitz <- bind_rows(sentiment_Gatsby, sentiment_Anthony, sentiment_Amory, sentiment_Dick_Diver) %>% mutate(total = rowSums(across(where(is.numeric)))) %>% mutate(across(anger:trust, .fns = ~ round((. / total) * 100)))

vis_ner_3 <- ggRadar(

  data = all_protag_fitz %>% select(-total),
  mapping = aes(color = title, facet = words),
  rescale = FALSE,
  interactive = TRUE,
  use.label = TRUE,
  size = 2,
  legend.position = "right"
)
vis_ner_3
```


```{r graphs_hemingway}
sentiment_Henry <- entities_matches_sentiment %>% filter(title == "A Farewell to Arms") %>%
  filter(words == "Henry") %>%
  group_by(title, words, kind) %>%
  summarise(across(anger:sentiment, sum)) %>% select(-positive, -negative, -sentiment, -kind)

sentiment_Robert <- entities_matches_sentiment %>% filter(title == "For Whom The Bell Tolls") %>%
  filter(words == "Robert Jordan") %>%
  group_by(title, words, kind) %>%
  summarise(across(anger:sentiment, sum)) %>% select(-positive, -negative, -sentiment, -kind)

sentiment_Jake <- entities_matches_sentiment %>% filter(title == "The Sun Also Rises") %>%
  filter(words == "Barnes") %>%
  group_by(title, words, kind) %>%
  summarise(across(anger:sentiment, sum)) %>% select(-positive, -negative, -sentiment, -kind)

sentiment_David <- entities_matches_sentiment %>% filter(title == "The Garden of Eden") %>%
  filter(words == "David") %>%
  group_by(title, words, kind) %>%
  summarise(across(anger:sentiment, sum)) %>% select(-positive, -negative, -sentiment, -kind)

all_protag_hemi <- bind_rows(sentiment_Henry, sentiment_Robert, sentiment_Jake, sentiment_David) %>% mutate(total = rowSums(across(where(is.numeric)))) %>% mutate(across(anger:trust, .fns = ~ round((. / total) * 100)))

vis_ner_4 <- ggRadar(

  data = all_protag_hemi %>% select(-total),
  mapping = aes(color = title, facet = words),
  rescale = FALSE,
  interactive = TRUE,
  use.label = TRUE,
  size = 2,
  legend.position = "right"
)
vis_ner_4

```

```{r exporting_tables}
save(vis_ner_1_int, vis_ner_2_int, vis_ner_3, vis_ner_4, file = "saved_ner.rda")
```
